---
title: "Assignment 1"
author: "glucksadam 300619334"
date: "July 2024"
fontsize: 10pt
output:
  pdf_document: null
  theme: united
  df_print: kable
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages

```{r packages,results='hide', collapse=TRUE, message=FALSE}
library(knitr)
library(MASS)
library(testthat)
```

<!-- Settings and Options -->
<!--  -  -->
<!--  - -->
<!--  -  -->

# Project Files

Assignment 1 project files are available at this [link](https://github.com/adamglux/A1DATA501.git).

Commit files are available at this [link](https://github.com/adamglux/A1DATA501/commits/master/)

# Shapiro-Wilk Test Statistic 

## Function 

This function follows the method:

$$W = \frac{(\sum^n_{i=1}a_ix_{(i)})^2}{\sum^n_{i=1}(x_i - \bar x)^2}$$

where $x_{(i)}$ is an ordered statistic, $\bar x$ is the sample mean, and $a_i$ is given by: 

$$(a_1,...,a_n) = \frac{m^TV^{-1}}{C}$$
$$C = (m^TV^{-1}V^{-1}m)^{1/2}$$

sources: 

[Wikipedia](https://en.wikipedia.org/wiki/Shapiro-Wilk_test)
other sources include: 

[Shapiro-Wilk Expanded Test](https://real-statistics.com/tests-normality-and-symmetry/statistical-tests-normality-symmetry/shapiro-wilk-expanded-test/), Charles Zaiontz 2014

[An Extension of Shapiro and Wilk's W Test for
Normality to Large Samples](https://www.jstor.org/stable/2347973?origin=crossref&seq=2), J.P. Royston

### SW_test Function

```{r function, echo=FALSE}
SW_test <- function(data, plot_qq = FALSE) { 
  # Input validation
  if (!is.numeric(data)) {
    stop("Input data must be numeric.")
  }
  
  if (length(data) < 3) {
    stop("Input data must have at least 3 values.")
  }
  
  if (any(is.na(data))) {
    stop("Input data contains NA values.")
  }
  
  if (any(is.infinite(data))) {
    stop("Input data contains infinite values.")
  }

  #Obtain sorted data, x_{(i)}
  n <- length(data)
  sorted_data <- sort(data)
  
  # compute expected values for a normal dist
  m <- qnorm((1:n - .375) / (n + .25))
  
  # compute a covariance matrix
  V <- matrix(0, n, n)
  for (i in 1:n) {
    for (j in 1:n) {
      V[i, j] <- min(i, j) / n - (i * j) / n^2
    }
  }

  # Compute inverse of V
  V_inv <- MASS::ginv(V)
  
  # Compute C
  #C <- sqrt(t(m) %*% V_inv %*% V_inv %*% m)
  
  # Compute weights a
  a <- t(m) %*% V_inv #V_inv %*% m
  C <- sqrt(sum(a^2))
  a <- a / C

  
  # Compute the mean of the data
  x_bar <- mean(data)

  # Compute the Shapiro-Wilk test statistic W
  W <- (sum(a * sorted_data)^2) / sum((sorted_data - mean(sorted_data))^2)
  
  # Plot QQ-plot if required
  if (plot_qq) {
    qqnorm(data)
    qqline(data, col = "blue", lwd = 2)
  }
  
  # Return W statistic, and QQ plot
 #return(list("Shapiro-Wilk W-statistic:" = W))
  return(list("W" = W))
  
}
```

## Test Data

### Create Test File

```{r echo=FALSE}
usethis::use_testthat()
```

### Testing in RMarkdown

```{r datasets}

### datasets

## generated data
data1 <- rnorm(300) 
data3 <- c(rnorm(50), NA)
data4 <- c(rnorm(50), Inf)
data5 <- c("a", "b", "c")
data6 <- c(1, 2)
data7 <- rbinom(300, 1000, prob = 0.7)

## imported data
milk <- read.csv("Data/Milk.csv") #available on assignment repo

```


```{r tests}
context("Testing SW_test context")

## invalid inputs 
test_that("function SW_test gives helpful errors", 
          {
            expect_error(SW_test(data3), "Input data contains NA values.")
            
            expect_error(SW_test(data4), "Input data contains infinite values.")
            
            expect_error(SW_test(data5), "Input data must be numeric.")
            
            expect_error(SW_test(data6), "Input data must have at least 3 values.")
            
            })

## valid inputs
test_that("function SW_test works with numeric dataframes", {
  
  expect_silent(SW_test(data1))
  
  expect_silent(SW_test(milk$Cost))
  
  expect_type(SW_test(data1), "list")
  
  expect_true(typeof(SW_test(data1)[[1]])=="double", TRUE)
})

```

### Testing using R file

**note**: the "R/test.R" file is a duplicate of the code above

```{r}
test_file("R/test.R")
```


# Sample Output 

```{r}
#basic output
SW_test(rnorm(100)) #sample size of 100 
SW_test(milk$Cost) #sample size of 12

#W from a non-normal distribution 
SW_test(data7)
#result is much smaller

## adding numbers to set
nums <- c(milk$Cost,runif(10, 2,6))
SW_test(nums)
```

A sample of 100 random normal has a W of 0.1052131. Datasets with smaller numbers, for example `milk` (n=12), and `nums` (n=5) have larger values for W. Adding more samples to `nums` brings down the value of W. It appears that low values of n are unstable. 



```{r output1, fig.height = 4, fig.width = 5, fig.align='center'}
SW_test(data2, plot_qq = T)
```


## As *n* samples increases

```{r output2, fig.height = 4, fig.width = 5, fig.align='center'}


#what happens to W as the size of the data increases? 
 w <- rep(NA,50)
for (i in 1:50) {
  data <- rnorm(10*i)
  dub <- SW_test(data)[[1]]
  w[i] <- dub[[1]]
}

plot(w, type="n")
lines(w)
```

As n increases, we get a smaller value for w, converging aroung 0.03. 

## compare QQ plots with increasing values of n

```{r output3, fig.height = 4, fig.width = 5, fig.align='center'}

par(mfrow = c(2,3))
SW_test(rnorm(10), plot_qq = T)
SW_test(rnorm(25), plot_qq = T)
SW_test(rnorm(50), plot_qq = T)
SW_test(rnorm(100), plot_qq = T)
SW_test(rnorm(200), plot_qq = T)
SW_test(rnorm(500), plot_qq = T)
```
As n samples of random normal increase, the quantities converge onto the QQ line. 

# Final Thoughts

The function `SW_test()` derived from Wikipedia and online sources output a test statistic, W, as well as a QQ-plot — when default parameter `FALSE` is changed to `TRUE`. These test statistics are sensitive to the number of samples. Here we see that samples derived from a random normal distribution have higher values than those derived from a binomial distribution.  

8 tests are run on the function, using two methods: code available in this document above, as well as a test file. Both use the same code. 4 Tests check error, and 4 tests check that the output is desired. All tests are passing, indicating that the correct errors are displayed, and the output is congruent with the functionality of the function.  

**Note:** Assignment 1 project files are available at this [link](https://github.com/adamglux/A1DATA501.git).


_____


\begin{center}
End of Assignment 1, DATA501

\end{center}

_____

## References

[Wikipedia](https://en.wikipedia.org/wiki/Shapiro-Wilk_test)
other sources include: 

[Shapiro-Wilk Expanded Test](https://real-statistics.com/tests-normality-and-symmetry/statistical-tests-normality-symmetry/shapiro-wilk-expanded-test/), Charles Zaiontz 2014

[An Extension of Shapiro and Wilk's W Test for
Normality to Large Samples](https://www.jstor.org/stable/2347973?origin=crossref&seq=2), J.P. Royston
